Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-09-07 docs/Manuscript/Manuscript.tex
make further revisions following discussion with adi

@@

Following the selection sources aligning with our review, we collected
characteristics of included studies (Section \ref{sec:availability}).
The features included bibliographic fields (date of publication,
journal, authors, title, keywords, DOI), study goals (objective,
generalizable knowledge, evaluation, clinical/medical domain), data sets
(data source, type of data, range of data, number of cases/incidences,
number of predictors/features), and methodological choices (types of
similarity measure, families of adaptation step/statistical model,
method(s) compared against, evaluation/performance/comparison measures
(\& their values), name given to modeling approach). We used these data
to detect and visualize trends amongst the included studies.

@@

The 25 included studies were analyzed based on 20 characteristics (see
Section \ref{sec:availability}), and we report and describe some
observations here (Table \ref{tab:synthesis}). The studies were
published in a variety of journals, with some of greater frequency,
though no single journal published more than 3. The journal
\emph{Evolving Systems} published 2 of the included studies,
\emph{Artificial Intelligence in Medicine} published 3, and
\emph{Hindawi - Journal of Healthcare Engineering} published 3.

@@

The included studies hypothesized, asserted, or assumed several benefits
of localized modeling specific to clinical and medical settings and
advantages over other modeling approaches (see Section
\ref{sec:availability}). The most common was that the restriction to
similar or relevant past cases would improve predictive performance for
the index case {[}11,15,16,26{]}. In particular, {[}14{]} hypothesized
and confirmed that the value of each past case would be positively
related to its similarity to the index case, an assumption built in to
the weighting schemes of other approaches. {[}12{]} made a different
case, that the fewer parametric assumptions and complications of a
CBR-style model would allow for greater accuracy. Additionally, several
investigators asserted that the use of localized cohorts befit the
clinical focus on the individual patient rather than the population,
without reference to model performance {[}21,23{]}.

@@

\hypertarget{method-development}{%
\subsection{Method development}\label{method-development}}

\textbf{Move part of this paragraph to Discussion.}

@@

@@

The study authors focused their recommendations and their own plans for
future work mostly on technical improvements and evaluations
{[}11,19{]}. Urged improvements included full or partial automation of
user-dependent steps such as predictor selection {[}11,19{]}, similarity
learning {[}11,17{]}, and parameter optimization {[}16,21{]}; extensions
to new data structures {[}24{]}, data types {[}25,26,31{]}, and
reasoning systems {[}28{]}; and the use of more advanced model
components to improve accuracy or efficiency {[}12,14,26,30{]}. Authors
also urged validations and independent evaluations using larger or more
comprehensive data sets {[}15,22,23,25{]}, using data aggregated from
multiple health systems {[}14,16,34,35{]}, and in other care settings or
disease contexts {[}21,30,34,35{]}. Less common were calls to strengthen
the connection between the methods and the users. Some authors urged the
incorporation of (human-derived) domain knowledge into the data or
models {[}17,19{]}. Others suggested new uses of their modeling
approaches: to measure feature importance {[}20{]}, to make models more
expressive {[}14{]}, and to combine information obtained both from
global and from localized models {[}15{]}. Most of this incremental work
was indeed carried out by later researchers whose output appears in our
sample.

@@

\hypertarget{trends}{%
\subsection{Trends}\label{trends}}

\textbf{Add 1--2 paragraphs.}

@@

The similarity learning approaches took a variety of forms. Several
studies used unsupervised similarity learning, for example the
Mahalanobis distance {[}12{]} and the Kolmogorov entropy-based distance
{[}22{]}. In particular, {[}19{]} defined similarity as a weighted sum
of differences in predictor values and used linear regression to
optimize the weights for the predictive accuracy of the cohort they
retrieve. Others used supervised learning: {[}21{]} proposed an
iterative algorithm to optimize a set of fuzzy inference rules used to
retrieve relevant cases and generate a prediction, using
back-propagation on the rules' parameters. Their method was used in
several later studies returned by our search, which are not discussed
here because they did not originate the technique. {[}28{]} explicitly
appeal to a supervised distance metric learning technique {[}43{]} to
obtain a similarity measure on their set of dermoscopy and confocal
images that most effectively separates malignant from benign melanoma
tumors. We will discuss the remaining uses of supervised similarity
learning shortly.

@@

@@

\hypertarget{data-and-code-availability}{%
\section{Data and code availability}\label{data-and-code-availability}}

@@

>>>>

\usepackage{lscape} \newcommand{\landscapebegin}{\begin{landscape}} \newcommand{\landscapeend}{\end{landscape}}

@@

Following the selection sources aligning with our review, we collected
characteristics of included studies (Section \ref{sec:data}). The
features included bibliographic fields (date of publication, journal,
authors, title, keywords, DOI), study goals (objective, generalizable
knowledge, evaluation, clinical/medical domain), data sets (data source,
type of data, range of data, number of cases/incidences, number of
predictors/features), and methodological choices (types of similarity
measure, families of adaptation step/statistical model, method(s)
compared against, evaluation/performance/comparison measures (\& their
values), name given to modeling approach). We used these data to detect
and visualize trends amongst the included studies.

@@

The 25 included studies were analyzed based on 20 characteristics (see
Section \ref{sec:data}), and we report and describe some observations
here (Table \ref{tab:synthesis}). The studies were published in a
variety of journals, with some of greater frequency, though no single
journal published more than 3. The journal \emph{Evolving Systems}
published 2 of the included studies, \emph{Artificial Intelligence in
Medicine} published 3, and \emph{Hindawi - Journal of Healthcare
Engineering} published 3.

@@

\label{sec:rationales}

The included studies hypothesized, asserted, or assumed several benefits
of localized modeling specific to clinical and medical settings and
advantages over other modeling approaches (see Section \ref{sec:data}).
The most common was that the restriction to similar or relevant past
cases would improve predictive performance for the index case
{[}11,15,16,26{]}. In particular, {[}14{]} hypothesized and confirmed
that the value of each past case would be positively related to its
similarity to the index case, an assumption built in to the weighting
schemes of other approaches. {[}12{]} made a different case, that the
fewer parametric assumptions and complications of a CBR-style model
would allow for greater accuracy. Additionally, several investigators
asserted that the use of localized cohorts befit the clinical focus on
the individual patient rather than the population, without reference to
model performance {[}21,23{]}.

@@

\hypertarget{challenges}{%
\subsection{Challenges}\label{challenges}}

@@

\hypertarget{performance-assessments}{%
\subsection{Performance assessments}\label{performance-assessments}}

\label{sec:performance}

Most included studies quantitatively compared the predictive performance
of their proposed method(s) to one or more comparators. What we took to
be the signature results are collated in Table \ref{tab:performance} in
the Appendix. Note that we exercised some judgment in classifying
methods as proposals and comparators, as in some cases all methods were
original but only some showcased main ideas. It would be impractical to
meta-analyze these numbers due to the great variety of settings,
problems, data types, and choices involved.

However, we do observe one clear pattern: Of the proposed methods that
most evidently outperformed their comparators (Kohonen + CBR {[}20{]},
TWNFI {[}18{]}, GSA {[}26{]}, CBR + rules (+ DML) {[}28{]}, Gaussian
process regression {[}30{]}, JITL-ELM {[}32{]}), all except the last are
hybrids of localized modeling (in some cases CBR) with other techniques,
often metric learning. Though {[}27{]} and {[}15{]} report exceptions,
the pattern suggests the importance of the similarity measure to the
retrieval step. Meanwhile, when proposed approaches targeted cohort
demarcation or choice of predictive model (statistical CBR {[}13{]};
individualized logistic regression, decision tree, and random forest
{[}14,16{]}), they did not consistently outperform comparators.

@@

\label{sec:needs}

The study authors focused their recommendations and their own plans for
future work mostly on technical improvements and evaluations
{[}11,19{]}. Urged improvements included full or partial automation of
user-dependent steps such as predictor selection {[}11,19{]}, similarity
learning {[}11,17{]}, and parameter optimization {[}16,21{]}; extensions
to new data structures {[}24{]}, data types {[}25,26,31{]}, and
reasoning systems {[}28{]}; and the use of more advanced model
components to improve accuracy or efficiency {[}12,14,26,30{]}. Authors
also urged validations and independent evaluations using larger or more
comprehensive data sets {[}15,22,23,25{]}, using data aggregated from
multiple health systems {[}14,16,34,35{]}, and in other care settings or
disease contexts {[}21,30,34,35{]}. Less common were calls to strengthen
the connection between the methods and the users. Some authors urged the
incorporation of (human-derived) domain knowledge into the data or
models {[}17,19{]}. Others suggested new uses of their modeling
approaches: to measure feature importance {[}20{]}, to make models more
expressive {[}14{]}, and to combine information obtained both from
global and from localized models {[}15{]}. Most of this incremental work
was indeed carried out in later investigations in our sample.

\hypertarget{public-availability}{%
\subsection{Public availability}\label{public-availability}}

\textbf{Discuss public availability of code or implementations.}

@@

\hypertarget{study-designs}{%
\subsection{Study designs}\label{study-designs}}

@@

The similarity learning approaches took a variety of forms. Several
studies used unsupervised similarity learning, for example the
Mahalanobis distance {[}12{]} and the Kolmogorov entropy-based distance
{[}22{]}. In particular, {[}19{]} defined similarity as a weighted sum
of differences in predictor values and used linear regression to
optimize the weights for the predictive accuracy of the cohort they
retrieve. Others used supervised learning: {[}21{]} proposed an
iterative algorithm to optimize a set of fuzzy inference rules used to
retrieve relevant cases and generate a prediction, using
back-propagation on the rules' parameters. Their method was used in
several later studies returned by our search, which are not discussed
here because they did not originate the technique. {[}28{]} explicitly
appeal to a supervised distance metric learning (DML) technique {[}43{]}
to obtain a similarity measure on their set of dermoscopy and confocal
images that most effectively separates malignant from benign melanoma
tumors. We will discuss the remaining uses of supervised similarity
learning shortly.

@@

\hypertarget{directions-and-expectations}{%
\subsection{Directions and
expectations}\label{directions-and-expectations}}

As discussed in Section \ref{sec:rationales}, most studies were premised
on the potential predictive value of localized models. As cautioned in
Section \ref{sec:performance}, not all experiments affirmed this
premise. Most of those that did involved using metric learning to
improve retrieval and sometimes iterative optimization of the metric and
of the localized model. Some studies showcased results using a variety
of such specializations {[}15,27,30{]}. However, because these advanced
implementations were mostly compared against their precursors or
commonplace alternatives, we cannot speak to their performance or any
trade-offs with respect to each other. This would be a worthy goal of
future work, though it might be limited by the lack of open-source code
or public implementations (Section \ref{sec:availability}).

While most studies identified outstanding technical needs (Section
\ref{sec:needs}), few emphasized the need to assess human-focused
qualities like user interface and user experience, interpretability,
meaningfulness, or trust. Most studies also identified interpretability
as an advantage of localized models, though none evaluated model
interpretability and few proposed new interpretative uses. Lack of trust
in ML tools is a long-recognized problem that direct intepretability of
model components could help alleviate, but this is more often assumed
than demonstrated. One valuable direction for future work would be to
compare the ability of non-technical investigators or users to draw
correct inferences from these and other interpretable methods and their
levels of confidence in those inferences.

We posit that an essential component of such ``soft'' performance goals
is the ability of the user community to exert some control over the
models. Given the impact on performance of the choice or optimization of
the similarity measure, an important target for user input would be the
importance of certain variables in the calculation of similarity, with
an understanding of how it can impact not only the performance of the
predictions but also the cohort retrieved for the model. For one
example, a user may want to minimize the weight of rare diseases in
medical history in order to retrieve a population with more such cases
in order to better measure their associated risk to the outcome. In
contrast, they may want to increase the weight of the indicating
diagnosis in order to allow fewer patients from similar but distinct
populations to influence the model. For another example, a user may want
to down-weight socioeconomic variables like race--ethnicity in order to
ensure a more diverse modeling cohort. Future purely quantitative work
could assess whether similarity-tuning can achieve these ends more
flexibly than strict inclusion/exclusion criteria.

@@

\hypertarget{data-and-code}{%
\section{Data and code}\label{data-and-code}}

\label{sec:data}

@@

\hypertarget{performance-evaluations-and-comparisons}{%
\subsection{Performance evaluations and
comparisons}\label{performance-evaluations-and-comparisons}}

\begin{landscape}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1224}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0714}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1633}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.2398}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0969}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.2092}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0969}}@{}}
\caption{\label{tab:performance}Evaluations of proposed methods and
comparisons to alternative methods. (See original studies for full names
and descriptions.)}\tabularnewline
Measure
Problem
Proposals
Comparators
Measure
Problem
Proposals
Comparators
{[}20{]} & Accuracy & Arthritis & Kohonen + CBR & 75.60\%\hspace{6em} &
Kohonen mapping & 54.70\%\hspace{6em} \\
& & & & \hspace{6em} & CBR & 60.40\%\hspace{6em} \\
& & & & \hspace{6em} & Quest & 50.90\%\hspace{6em} \\
& & & & \hspace{6em} & backpropagation & 54.70\%\hspace{6em} \\
{[}13{]} & & Dermatology & Statistical CBR & 96.57\%\hspace{6em} & CBR &
95.43\%\hspace{6em} \\
& & & & \hspace{6em} & C5.0 & 93.71\%\hspace{6em} \\
& & & & \hspace{6em} & CART & 92.00\%\hspace{6em} \\
& & & & \hspace{6em} & LR & 91.14\%\hspace{6em} \\
& & & & \hspace{6em} & NN & 88.29\%\hspace{6em} \\
& & Heart & Statistical CBR & 83.70\%\hspace{6em} & LR &
84.44\%\hspace{6em} \\
& & & & \hspace{6em} & NN & 82.22\%\hspace{6em} \\
& & & & \hspace{6em} & CBR & 81.85\%\hspace{6em} \\
& & & & \hspace{6em} & CART & 76.67\%\hspace{6em} \\
& & & & \hspace{6em} & C5.0 & 76.30\%\hspace{6em} \\
& & Breast & Statistical CBR & 96.96\%\hspace{6em} & NN &
97.50\%\hspace{6em} \\
& & & & \hspace{6em} & CBR & 96.61\%\hspace{6em} \\
& & & & \hspace{6em} & LR & 96.61\%\hspace{6em} \\
& & & & \hspace{6em} & C5.0 & 94.29\%\hspace{6em} \\
& & & & \hspace{6em} & CART & 92.14\%\hspace{6em} \\
& & Diabetes & Statistical CBR & 76.32\%\hspace{6em} & LR &
77.11\%\hspace{6em} \\
& & & & \hspace{6em} & CBR & 73.16\%\hspace{6em} \\
& & & & \hspace{6em} & C5.0 & 73.15\%\hspace{6em} \\
& & & & \hspace{6em} & CART & 72.89\%\hspace{6em} \\
& & & & \hspace{6em} & NN & 65.39\%\hspace{6em} \\
& & Liver & Statistical CBR & 66.76\%\hspace{6em} & CART &
67.94\%\hspace{6em} \\
& & & & \hspace{6em} & LR & 67.35\%\hspace{6em} \\
& & & & \hspace{6em} & C5.0 & 66.47\%\hspace{6em} \\
& & & & \hspace{6em} & CBR & 60.88\%\hspace{6em} \\
& & & & \hspace{6em} & NN & 55.59\%\hspace{6em} \\
{[}21{]} & RMSE & GFR & TWNFI & 7.08\hspace{6em} & MDRD &
7.74\hspace{6em} \\
& & & & \hspace{6em} & MLP & 8.38\hspace{6em} \\
& & & & \hspace{6em} & ANFIS & 7.40\hspace{6em} \\
& & & & \hspace{6em} & DENFIS & 7.22\hspace{6em} \\
& & & & \hspace{6em} & TNFI & 7.28\hspace{6em} \\
{[}22{]} & AUROC & mammography mass region & CBR & 0.885\hspace{6em} &
DT & 0.872\hspace{6em} \\
& & & & \hspace{6em} & ANN & 0.882\hspace{6em} \\
{[}23{]} & Accuracy & Return-to-work & CBR & 62.5\%\hspace{6em} & LR &
71.2\%\hspace{6em} \\
{[}18{]} & & Colon cancer diagnosis & TWNFI & 91.9\%\hspace{6em} & MLR
(Personalized) & 82.3\%\hspace{6em} \\
& & & & \hspace{6em} & SVM (Personalized) & 90.3\%\hspace{6em} \\
& & & & \hspace{6em} & WKNN & 90.3\%\hspace{6em} \\
& & Crohn's disease risk & TWNFI & 82.45\%\hspace{6em} & &
\hspace{6em} \\
{[}26{]} & & Colon cancer & knnGSA & 85.48\%\hspace{6em} & SVM &
82.14\%\hspace{6em} \\
& & & svmGSA & 87.10\%\hspace{6em} & ECF & 72.30\%\hspace{6em} \\
& & & & \hspace{6em} & KNN & 82.14\%\hspace{6em} \\
& & & & \hspace{6em} & WKNN & 82.14\%\hspace{6em} \\
& & Leukemia & knnGSA & 97.22\%\hspace{6em} & SVM &
95.83\%\hspace{6em} \\
& & & svmGSA & 97.22\%\hspace{6em} & ECF & 94.44\%\hspace{6em} \\
& & & & \hspace{6em} & KNN & 94.44\%\hspace{6em} \\
& & & & \hspace{6em} & WKNN & 94.44\%\hspace{6em} \\
& & Lymphoma & knnGSA & 94.81\%\hspace{6em} & SVM &
93.51\%\hspace{6em} \\
& & & svmGSA & 94.81\%\hspace{6em} & ECF & 92.21\%\hspace{6em} \\
& & & & \hspace{6em} & KNN & 93.51\%\hspace{6em} \\
& & & & \hspace{6em} & WKNN & 93.51\%\hspace{6em} \\
& & Lung cancer & knnGSA & 98.34\%\hspace{6em} & SVM &
95.31\%\hspace{6em} \\
& & & svmGSA & 98.90\%\hspace{6em} & ECF & 92.87\%\hspace{6em} \\
& & & & \hspace{6em} & KNN & 96.60\%\hspace{6em} \\
& & & & \hspace{6em} & WKNN & 95.50\%\hspace{6em} \\
{[}12{]} & Relative IPEC & Graft survival & MKNN & 0.973\hspace{6em} &
RSF & 0.957\hspace{6em} \\
{[}27{]} & AUROC & Waitlist registration, complete & CBR + Wj &
89.8\%\hspace{6em} & LR & 92.0\%\hspace{6em} \\
& & & CBR + Wk & 90.7\%\hspace{6em} & Standalone CBR &
90.4\%\hspace{6em} \\
& & & CBR + Wj + Wk & 87.4\%\hspace{6em} & & \hspace{6em} \\
& & Waitlist registration, stepwise & CBR + Wj & 82.7\%\hspace{6em} & LR
& 92.1\%\hspace{6em} \\
& & & CBR + Wk & 86.2\%\hspace{6em} & Standalone CBR &
91.4\%\hspace{6em} \\
& & & CBR + Wj + Wk & 88.5\%\hspace{6em} & & \hspace{6em} \\
{[}28{]} & Accuracy & Nevus & Collaborative + rules & 98\%\hspace{6em} &
Dermoscopy CBR & 92\%\hspace{6em} \\
& & & Collaborative + rules + DML & 100\%\hspace{6em} & Confocal CBR &
96\%\hspace{6em} \\
& & & & \hspace{6em} & Collaborative & 95\%\hspace{6em} \\
{[}15{]} & AUROC & Diabetes onset & Personalized LR
(LSML+FeatureFiltering) & 0.624\hspace{6em} & KNN & 0.617\hspace{6em} \\
& & & Personalized LR (LSML) & 0.619\hspace{6em} & Global LR &
0.611\hspace{6em} \\
& & & Personalized LR (Euclidean) & 0.614\hspace{6em} & &
\hspace{6em} \\
& & & Personalized LR (Random) & 0.602\hspace{6em} & & \hspace{6em} \\
{[}14{]} & & 30-day mortality & Individualized LR & 0.830\hspace{6em} &
Individualized DC & 0.797\hspace{6em} \\
& & & Individualized DT & 0.753\hspace{6em} & & \hspace{6em} \\
{[}16{]} & & & Individualized LR & 0.824\hspace{6em} & Individualized DC
& 0.801\hspace{6em} \\
& & & Individualized DT & 0.779\hspace{6em} & & \hspace{6em} \\
& & & Individualized RF & 0.839\hspace{6em} & & \hspace{6em} \\
& & & Individualized CSRF & 0.832\hspace{6em} & & \hspace{6em} \\
{[}31{]} & Accuracy & J13/pneumonia & Case-based NN & 81.6\%\hspace{6em}
& & \hspace{6em} \\
& & K80.1/gallbladder & Case-based NN & 76.7\%\hspace{6em} & &
\hspace{6em} \\
& & H25.1/age-related cataract & Case-based NN & 94.9\%\hspace{6em} & &
\hspace{6em} \\
& & H26.2/complicated cataract & Case-based NN & 91.4\%\hspace{6em} & &
\hspace{6em} \\
& & 167.4/encephalopathy & Case-based NN & 72.4\%\hspace{6em} & &
\hspace{6em} \\
& & 167.9/CBD & Case-based NN & 75.4\%\hspace{6em} & & \hspace{6em} \\
& & N20.1/ureter & Case-based NN & 58.7\%\hspace{6em} & &
\hspace{6em} \\
{[}30{]} & AUROC & Cognitive impairment & Gaussian process regression
with custom kernel & 0.92\hspace{6em} & linear regression without
regularization & 0.81\hspace{6em} \\
& & & & \hspace{6em} & LASSO regression & 0.85\hspace{6em} \\
& & & & \hspace{6em} & ridge regression & 0.87\hspace{6em} \\
& & & & \hspace{6em} & RF & 0.90\hspace{6em} \\
& & & & \hspace{6em} & gradient boosting regression tree &
0.91\hspace{6em} \\
& & & & \hspace{6em} & XGBoost & 0.91\hspace{6em} \\
& & & & \hspace{6em} & kernel support vector regressor &
0.92\hspace{6em} \\
& & Parkinson's disease & Gaussian process regression with custom kernel
& 0.875\hspace{6em} & linear regression without regularization &
0.820\hspace{6em} \\
& & & & \hspace{6em} & ridge regression & 0.798\hspace{6em} \\
& & & & \hspace{6em} & RF & 0.734\hspace{6em} \\
& & & & \hspace{6em} & gradient boosting regression tree &
0.825\hspace{6em} \\
& & & & \hspace{6em} & kernel support vector regressor &
0.778\hspace{6em} \\
{[}17{]} & & Diabetes & Personalized RF & 0.90\hspace{6em} & &
\hspace{6em} \\
& & & Personalized KNN & 0.82\hspace{6em} & & \hspace{6em} \\
& & & Personalized LR & 0.89\hspace{6em} & & \hspace{6em} \\
{[}32{]} & & Discharge at 10 days & One-class JITL-ELM &
0.8510\hspace{6em} & ELM & 0.4973\hspace{6em} \\
& & & & \hspace{6em} & JITL-ELM & 0.2014\hspace{6em} \\
& & & & \hspace{6em} & One-class ELM & 0.7588\hspace{6em} \\
& & & & \hspace{6em} & One-class SVM & 0.4647\hspace{6em} \\

\end{landscape}
